Does at-issueness predict projection? Further investigations of the Gradient Projection Principle

### 2022-12-06 writing/investigating-GPP.tex
updated paper

>>>>

\subsection{Factive, semi-factive, etc. references}

{\bf IDEA:} Look for factive and semi-factive (cognitive and emotive) predicates in CommitmentBank, VerbVeridicality, and MegaVeridicality

\item \citealt[65]{karttunen71b}

``Finally, there are at least two distinct types of factive verbs, factives and semi-factives, that differ from each other in contexts that involve certain modal operators, although they are not distinct in simple negative or affirmative assertions.''

\item \citealt[63ff.]{karttunen71b}

{\tt Whatever a sentence with a factive predicate presupposes, the presupposition ought to remain no matter whether the
main sentence is a negative assertion, an interrogative sentence, or the antecedent of a conditional construction.

For most of the verbs listed in (1) this is true. It is clear that all the examples in (22) presuppose that John had not told the truth.

(22) John didn't regret / realize / discover that he had not told the truth.

The only circumstance in which the examples would not carry along a commitment to the view that John had not told the truth is a case where the assertions in (22) are uttered as an emphatic denial of somebody else's previous assertion. That is, one might say something like (23)

(23) John DIDN'T regret that he had not told the truth. How could he have done that when he knew that what he had said was true?

If we leave examples of this sort out of consideration, there are no problems with negative assertions.7 All of the verbs listed in (1) meet this test for factivity. However, it seems that, in questions, some of the verbs behave differ- ently. Consider the examples in (24).

(24) Did you regret / realize / discover that you had not told the truth?

Most of my informants agree that with regret as the main verb, and possibly also with realize, the question in (24) commits the speaker to the view that the addressee has not told the truth. However, (24c) can also be understood as a sincere request for information. The speaker is not sure about the truth of the complement and is prepared to accept the address- ee's discovery as a fact. In addition to the verb discover, verbs like find out, and see differ from the other factive verbs in this respect. In questions, they permit both a factive and non-factive interpretation.

In conditionals, the verbs that lose their factivity certainly also include realize. Consider the examples in
(25).

(25) If I regret / realize / discover later later that I have not told the truth, I will confess it to everyone.

In uttering (25b) and (25c) one is not making a confession. With realize and discover as predicate, one only admits that
there is a possibility that one has not told the truth, but one does not concede anything more than that. With regret as the main verb, the first clause of (25) contains an admission that the complement is true.

As I suggested earlier in connection with the examples in (13) and (14), it seems that a conditional sentence conversationally implies that, in the view of the speaker, it is at least possible for the antecedent to turn out to be true. By this analysis, the examples in (25) conversa- tionally imply the corresponding sentences in (26).
(26) It is possible that I will regret / realize / discover later that I have not told the truth.

Given (11'), the stronger pair of meaning postulates for regret, it is clear that from (26a) one can infer the truth of the complement. Since there is no such necessary relation ship in (26b) and (26c), the corresponding stronger meaning
postulates for realize and discover would fail. From the fact that it is possible that I may discover something I cannot conclude that this something is in fact the case.
However, for regret, forget, resent, and all the factives that take sentential subjects this kind of reasoning is valid. The meaning postulates for discover, realize, find out, see, notice, and other similar verbs - call them 'semi- factives' - have to be of the weaker form given in (11).
}

\item \citealt[479f.]{hooper-thompson1973} follow Karttunen 1971.

\item \citealt{djaerv-etal2018}

\item Footnote 2: Here we follow H\&T in our use of `factive' and `semifactive'; these two classes are now commonly referred to as `emotive factives' and `cognitive factives' respectively.

\item p.6: 
{\tt Class D predicates e.g. resent, regret, be sorry, be surprised, be interesting. These are the (emotive) factive predicates. Given H\&T?s assumptions about the relation between factivity, presupposition, and assertion, the use of these verbs entails that the com- plement of these verbs is presupposed, and cannot be asserted.

Class E predicates e.g. realize, learn, discover, know.3 This group constitutes the semi- factives, which presuppose/entail the truth of their complements only in some environments ? in particular, this presupposition can be lost in questions and con- ditionals (Kiparsky \& Kiparsky 1970).}

\item Footnote 3: Whether know should be grouped together with the semifactives is highly contentious; here we take no position on this.

\item p.24: EV2 was most acceptable for semifactives and communicative assertives. Compared with assertive predicates that explicitly convey a communicative act, epis- temic (or cognitive) assertives yielded much lower ratings for EV2. And finally, EV2 was least acceptable under factives.

\item Dj\"arv Factivity, Assertion, and Clausal Definiteness

p.170: ``As many readers will no doubt be aware, Karttunen (1971) referred to cognitive factives like discover
and realize as ?semi-factives?; a term adopted by Hooper & Thompson (1973) and subsequent work on
MCP, and generally contrasted with ?true? (emotive) factives like resent and regret. These terms suggest
that factives like discover are therefore somehow ?less factive? than factives like resent. This, however,
is not what Karttunen?s (1971) distinction refers to. Rather, the distinction has to do with the relative sensitivity of the presupposition to entailment-cancelling operators like modals and negation, in contexts
that establish speaker-ignorance with respect to p (see for instance Heim 1983; Simons 2001).''
